Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-13 paper-new/ordering.tex
changes to term precision syntax + semantics

The rules for term precision come in two forms. We first have the \emph{congruence} rules,
one for each term constructor. These assert that the term constructors respect term precision.
The congruence rules are as follows:

@@

  \inferrule*[right = Bind]
    {\etmprec {\gamlt} {M} {M'} {c} \and 
     \etmprec {\gamlt, x : c} {N} {N'} {d} }
    {\etmprec {\gamlt} {\bind {x} {M} {N}} {\bind {x} {M'} {N'}} {d}}
\end{mathpar}

We then have additional equational axioms, including $\beta$ and $\eta$ laws, and
rules characterizing upcasts as least upper bounds, and downcasts as greatest lower bounds.
For the sake of familiarity, we formulate the cast rules using arbitrary casts; later we
will state the analogous versions for the version of the calculus without arbitrary casts.

@@

\begin{mathpar}
  \inferrule*[right = $\err$]
    {\phasty {\Gamma} M B }
    {\etmprec {\Gamma} {\err_B} M B}

@@

  \inferrule*[right = DnR]
    { c : A \ltdyn B \and d : B \ltdyn C \and
      \etmprec {\gamlt} {M} {N} {c \circ d} } 
    { \etmprec {\gamlt} {M} {\dn {B} {C} N} {c} }
\end{mathpar}

% TODO explain the least upper bound/greatest lower bound rules
The rules UpR, UpL, DnL, and DnR were introduced in \cite{new-licata18} as a means
of cleanly axiomatizing the intended behavior of casts in a way that
doesn't depend on the specific constructs of the language.
Intuitively, rule UpR says that the upcast of $M$ is an upper bound for $M$
in that $M$ may error more, and UpL says that the upcast is the \emph{least}
such upper bound, in that it errors more than any other upper bound for $M$.
Conversely, DnL says that the downcast of $M$ is a lower bound, and DnR says
that it is the \emph{greatest} lower bound.
% These rules provide a clean axiomatization of the behavior of casts that doesn't
% depend on the specific constructs of the language.

@@

If $c$ is inj-arr, then when we downcast $M$ from $dyn$ to $\dyntodyn$,
semantically this will involve a $\theta$ because the value of type $dyn$
in the semantics will contain a \emph{later} function $\tilde{f}$.
Thus, in order for the right-hand side to be related to the downcast,
we need to insert a delay on the right.
%
The need for delays affects the cast rules involving upcasts as well, because
the upcast for functions involves a downcast on the domain:

@@

\subsubsection{Perturbations}

@@

>>>>

The rules for term precision come in two forms. We first have the \emph{congruence} rules,
one for each term constructor. These assert that the term constructors respect term precision.
The congruence rules are shown in Figure \ref{fig:term-prec-congruence-rules}.

\begin{figure*}
\begin{mathpar}

@@

  \inferrule*[right = Bind]
    {\etmprec {\gamlt} {M} {M'} {c} \and 
     \etmprec {\gamlt, x : c} {N} {N'} {d} }
    {\etmprec {\gamlt} {\bind {x} {M} {N}} {\bind {x} {M'} {N'}} {d}}
\end{mathpar}
\caption{Term precision: congruence rules.}
\label{fig:term-prec-congruence-rules}
\end{figure*}

We then have additional equational axioms as shown in Figure \ref{fig:term-prec-additional-rules}.
In particular, we have a rule stating that the error term is the bottom element,
$\beta$ and $\eta$ laws as order-equivalences, and lastly rules characterizing upcasts
as least upper bounds, and downcasts as greatest lower bounds.
% For the sake of familiarity, we formulate the cast rules using arbitrary casts; later we
% will state the analogous versions for the version of the calculus without arbitrary casts.

@@

\begin{figure*}
\begin{mathpar}
  \inferrule*[right = $\err$]
    {\phasty {\Gamma} M B }
    {\etmprec {\Gamma} {\err_B} M B}

@@

  \inferrule*[right = DnR]
    { c : A \ltdyn B \and d : B \ltdyn C \and
      \etmprec {\gamlt} {M} {N} {c \circ d} } 
    { \etmprec {\gamlt} {M} {\dn {B} {C} N} {c} }
\end{mathpar}
\caption{Term precision: additional rules.}
\label{fig:term-prec-additional-rules}
\end{figure*}

% TODO explain the least upper bound/greatest lower bound rules
The rules \textsc{UpR}, \textsc{UpL}, \textsc{DnL}, and \textsc{DnR}
were introduced in \cite{new-licata18} as a means
of cleanly axiomatizing the intended behavior of casts in a way that
doesn't depend on the specific constructs of the language.
Intuitively, rule \textsc{UpR} says that the upcast of $M$ is an upper bound for $M$
in that $M$ may error more, and \textsc{UpL} says that the upcast is the \emph{least}
such upper bound, in that it errors more than any other upper bound for $M$.
Conversely, \textsc{DnL} says that the downcast of $M$ is a lower bound, and \textsc{DnR} says
that it is the \emph{greatest} lower bound.
% These rules provide a clean axiomatization of the behavior of casts that doesn't
% depend on the specific constructs of the language.

@@

If $c$ is inj-arr, then when we downcast $M$ from $dyn$ to $\dyntodyn$,
semantically this will involve a $\theta$ because the value of type $dyn$
in the semantics will contain a \emph{later} function $\tilde{f}$.
Thus, in order for the right-hand side to be related to the downcast,
the right-hand side must be of the form $\theta(\dots)$, and the argument to
$\theta$ on the LHS must be related to the argument of $\theta$ on the RHS.
We can accomplish this by inserting a ``delay" on the right, i.e., wrapping the
RHS in $\delta = \theta \circ \nxt$.
% Maybe show the correct form of the rule in the special case $c = inj-arr$.
%
The need for delays affects the cast rules involving upcasts as well, because
the upcast for functions involves a downcast on the domain:

@@

The specific manner in which these delays must be inserted for a given type precision
derivation $c$ is best examined in a more abstract setting, as it is independent of the
particular concrete model under consideration.
Thus, we postpone this discussion to the section on the abstract intensional categorical models of
gradual typing (Section \ref{sec:abstract-intensional-models}).

\begin{comment}
\subsubsection{Perturbations}

@@

\end{comment}
